---
title: "More Data Visualization"
author: "Dominguez Center for Data Science"
date: "May 28, 2026"
format: pdf
execute:
  warning: false
  message: false
editor: visual
editor_options: 
  chunk_output_type: console
---

# Questions Round-up

# Load Packages

The packages we need for our explorations today (`readr` for reading in data and `ggplot2` for graphing data) are part of a popular suite of packages called the `tidyverse`.

```{r}
library(tidyverse)
library(ggrepel)
```

# Data Background

In 2013, the government decided to make data about colleges more accessible so that students and parents could more easily compare schools. These data are called the ["College Scorecard" data](https://collegescorecard.ed.gov/) and the 2024 dataset contains 3,305 variables on 6,484 universities in the US!

I have filtered that 2024 dataset to only include schools which confer majority baccalaureate degrees and where the majority of those degrees are in the arts and sciences based on the Carnegie Classification system. In other words, I filtered the data down to the schools which are "similar" to Bucknell (including Bucknell itself) and picked out some variables for us to explore.

Let's use these data to start exploring and visualizing with `R`!

## Data Dictionary

Below are the code names and descriptions of the variables in our dataset.

-   `UNITID`: Unique identifier

-   `INSTNM`: Name of institution

-   `CITY`: City

-   `STABBR`: State

-   `HIGHDEG`: Highest degree awarded (0 = Non-degree grants, 1 = Certificate degree, 2 = Associate degree, 3 = Bachelor's degree, 4 = Graduate degree)

-   `HIGHDEG_CAT`: Highest degree awarded (Bachelor's degree, Graduate degree)

-   `PREDDEG`: Predominant undergraduate degree awarded (0 = Not classified, 1 = Predominantly certificate-degree granting, 2 = Predominantly associate's-degree granting, 3 = Predominantly bachelor's-degree granting, 4 = Entirely graduate-degree granting)

-   `CONTROL`: Ownership (1 = Public, 2 = Private non-profit, 3 = Private for-profit)

-   `CONTROL_CAT` = Ownership (Public, Private non-profit, Private for-profit)

-   `HBCU`: Flag for Historically Black College and University

-   `TUITFTE`: Net tuition revenue per full-time equivalent student

-   `AVGFACSAL`: Average faculty salary

-   `ADM_RATE`: Admission rate

-   `SATVR75`: 75th percentile of SAT scores at the institution (critical reading)

-   `SATMT75`: 75th percentile of SAT scores at the institution (math)

-   `ACTCM75`: 75th percentile of the ACT cumulative score

-   `COSTT4_A`: The average annual total cost of attendance, including tuition and fees, books and supplies, and living expenses for all full-time, first-time, degree/certificate-seeking undergraduates who receive Title IV aid.

-   `NPT4_PRIV`: The average annual total cost of attendance (CostT4_A, CostT4_P), including tuition and fees, books and supplies, and living expenses, minus the average grant/scholarship aid

-   `UGDS`: Enrollment of undergraduate certificate/degree-seeking students

-   `UG25ABV`: Percentage of undergraduates aged 25 and above

-   `PCTFLOAN_DCS`: Percentage of degree/certificate-seeking undergraduate students awarded a federal loan

-   `PCTPELL_DCS`: Percentage of degree/certificate-seeking undergraduate students awarded a Pell Grant

-   `DEBT_MDN`: The median original amount of the loan principal upon entering repayment

-   `C100_4`: Completion rate for first-time, full-time students at four-year institutions (100% of expected time to completion)

-   `RET_FT4`: First-time, full-time student retention rate at four-year institutions

-   `MD_EARN_WNE_5YR`: Median earnings of graduates working and not enrolled 5 years after completing

-   `Location`: Whether or not the institution is located in Pennsylvania or not

# Load the Data

Run the following code to load and inspect the data.

```{r}
# Load the data
colleges <- read_csv("data/ccbasic21.csv") 


# Wrangling data (will discuss next time!)
colleges <- colleges %>%
  mutate(Location = if_else(STABBR == "PA", "PA", "Not PA"),
         CONTROL_CAT = case_match(CONTROL, 
                                  1 ~ "Public",
                                  2 ~ "Private non-profit"),
         HIGHDEG_CAT = case_match(HIGHDEG,
                                  3 ~ "Bachelor's degree",
                                  4 ~ "Graduate degree"))
  

# Inspect the data
glimpse(colleges)
```

# Reminder: `ggplot` Structure

**Guiding Principle**: We will map variables from the **data** to the **aes**thetic attributes (e.g. location, size, shape, color) of **geom**etric objects (e.g. points, lines, bars).

To create a ggplot, we should ask ourselves the following questions:

-   What is the name of the dataset?
-   What is the geom (shape)?
-   What are the variables?
-   How should we map the variables to the geom?

# Moving Beyond the Scatterplot

## `geom_histogram()`

Let's create a histogram of `ADM_RATE`. Let's also

-   Play around with the number of bins.
-   Make the bars and outline of the bars fun colors!

```{r}

```

## `geom_boxplot()`

Let's create boxplots of `ADM_RATE` by whether or not a school is private or public (`CONTROL`).

```{r}

# Or recode CONTROL

```

Let's add `HIGHDEG_CAT` to our boxplots by coloring the boxes based on the categories of `HIGHDEG_CAT.`

```{r}

```

## `geom_violin()`

Let's now create violin plots of `ADM_RATE` by whether or not a school is private or public (`CONTROL`).

```{r}


# Add the data points to the plot (with jitter)

```

Compare and contrast the information provided by the boxplot and the violin plot.

## `geom_bar()`

Let's create bar graphs that compare the number of public and private schools by highest degree awarded.

```{r}
# Graph of counts


# Graph of dodged counts


# Graph of proportions
       
```

# `fill` versus `color`

Here's a graph from yesterday where we used `scale_color_manual()`:

```{r}
ggplot(data = colleges,
       mapping = aes(x = ADM_RATE, 
                     y = COSTT4_A,
                     color = Location)) +
  geom_point(size = 2, alpha = .5) +
  scale_color_manual(values = c("slateblue", "orange"))
```

Let's try to change the colors in our plot:

```{r}

```

Look in your `aes()` and see if you are using `color` or `fill` before deciding the `scale_XXX()`.

# Resources for Learning More about Graphing with `ggplot2`

-   Modern Dive's chapter on [Data Visualization](https://moderndive.com/2-viz.html)
-   R for Data Science's chapter on [Data Visualization](https://r4ds.had.co.nz/data-visualisation.html)
-   `ggplot2` cheatsheet: <https://rstudio.github.io/cheatsheets/data-visualization.pdf>
